---
title: "Using the xrfr Package"
output: rmarkdown::html_vignette
author: Anna Grytaas
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Using the xrfr Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This package was created for use at the University of Bergen to convert data from an X-ray fluorescence machine from kilo-counts per second to micromolar. It requires that your files are constructed in a very specific way, as explained in this vignette, so is meant to be used with instructions. 

The functions included in this package are:

* **`importxrf()`**: This function joins the raw data file with the file containing project information like where each sample was taken and how much water was filtered. The output of this function is a dataframe. 
* **`convertxrf()`**: This function takes the dataframe created by `importxrf()` and the file containing information like crystal drift and detection limits for each element. The output of this function is a long dataframe where the counts from the machine have been converted from kcps to µM. 
* **`widen()`**: This function takes the CSV file from when you save the dataframe created with `convertxrf()` and makes it wider, so each element has its own column. 
* **`widen_above()`**: This function both widens your data like `widen()` does, but also excludes any concentrations that are below the detection limits. 
* **`widen_means()`**: This function both widens your data like `widen()` does, but also calculates mean concentration based on one or two factors. 
* **`importdata()`**: This is a function that is used inside the `importxrf()` function to import your raw data file. *You will not need to use this yourself unless there is an error with* `importxrf()` *and you are looking for the cause of the error.*
* **`importinfo()`**: This is a function that is used inside the `importxrf()` function to import your project information file. *You will not need to use this yourself unless there is an error with* `importxrf()` *and you are looking for the cause of the error.*
* **`importsetup()`**: This is a function that is used inside the `convertxrf()` function to import your file containing information like crystal drift and detection limits. *You will not need to use this yourself unless there is an error with* `convertxrf()` *and you are looking for the cause of the error.*

## The files

There are three files you need for your analysis: a file with **raw data** from the XRF machine. You also need a file with **information about the samples and project**, like which station the samples are from, whether they are blank samples, and the volume of the filtrated water. Lastly, you need a file containing **information about the machine, filters and elements**, like crystal drift and calibrated constants. 

### The raw data file

This file is created by the XRF machine when you export your data from it. You should not edit this file unless there are errors in the sample names. This file is a .txt file using comma (,) as a decimal mark and tabs to separate columns. It will contain a column named **"Sample"**, a column named **"Date"**, and several columns with results from the machine's analysis. The ones we are interested in in this case are those with "(Int)" behind the element. 

### The project information file

This file should contain all relevant information that you wish to include about your data that is not from the XRF machine. It must be saved as an Excel file (.xlsx or .xls). There are several columns that must be included for the functions in this package to work:

* **"Sample"**: The name of your sample. Must be the same as in the raw data file. Any samples that do not match exactly will not be included in the created file. 
* **"Filter_type"**: The type of filter used when filtering. Can be PC, ANO, or GFF. This column must be filled out even if the same filter type was used on all samples because R needs to know which detection limits and calibrated constants to use. 
* **"Filter_size"**: The pore size of the filter used. This column can be left empty if the size was the same in all filters, but the column must exist. 
* **"Filter_box_nr"**: Let's R know which box the filter came from so it knows which blank samples to apply when finding the net count in this sample. 
* **Filter_blank**: In this column you will need to write "blank" for all your blank samples.
* **Volume**: The volume of water that was filtered. 

You may include as many more columns as you want, for example dates, depth, or station, but these columns all need to be present in your project information file. 

### The "setup" file

This file must contain all the relevant information related to the XRF machine, filters, and elements. It must be saved as an Excel file (.xlsx or .xls). This is a file that will not change between experiments unless the crystal drift has been measured again. The "setup" file must contain these columns:

* **Element**: All the elements measured by the machine. 
* **MolarW**: The molar weight of each element. 
* **PC**, **ANO**, and **GFF**: The calibrated constant for each element for each of the three filter types. Must be in this order. 
* **DL_PC**, **DL_ANO**, and **DL_GFF**: The detection limit of each element for each of the three filter types. Must be in this order. 
* **Drift_2008**, **Drift_2010**, etc.: The crystal drift each year it was measured. 2008 was the first year this was measured on the XRF machine at UiB, so this is used during the calculations. 

## The project file

This is a file that is generated once you have performed the functions `importxrf()` and `convertxrf()` and saved the dataframe (by using `write_csv()`, a function in the readr package). It must be a CSV file. This file will be used for further analysis and/or to create graphics using R/Rstudio etc. 

## The functions in the xrfr package

This section explains what each function does, and which arguments you need to define to make it work. The examples given show real data from an experiment at UiB. The raw files are available [here](https://github.com/agryt/xrfr/tree/master/inst/extdata). 

### importxrf()

This function will import and rearrange your files with raw data and project information and then join them into one dataframe. To use this function there are two arguments you must define:

* **datapath**: This is the name of your raw data file from the XRF machine. The file must be a .txt file. 
* **infopath**: This is the name of your project information file. The file must be an Excel file, either .xlsx or .xls. 

Here is a possible input:

```{r importxrf, eval = FALSE}

df <- importxrf(datapath = "xrf_rawdata.txt", infopath = "xrf_projectinfo.xlsx")

```

This code will give a dataframe that looks something like [this](https://github.com/agryt/xrfr/blob/master/inst/extdata/project_midway.csv) (link to a whole data file showing some real samples - the "extdata" directory there also contains the raw files used in this explanation). The table shows the first rows and columns of the linked data file:

|Sample|Date.x|C|N|O|Na|Mg|Al|Si|P|S|Cl|...|
|:----|:------|:-|:-|:-|:-|:-|:-|:-|:-|:-|:-|:-|
|COM001|04.12.2019 10:16:35|58.0957|0.438|2.9889|0.4874|1.796|32.3646|5.6097|0.3945|1.0494|1.0383|...|
|COM002|04.12.2019 10:28:50|54.2361|0.4313|3.3708|1.1547|4.3948|37.2398|17.5392|0.5557|1.538|1.8231|...|
|COM003|04.12.2019 10:41:07|59.0261|0.4458|2.8878|0.4234|1.2589|33.6267|2.6512|0.3778|0.9904|1.2301|...|
|...|...|...|...|...|...|...|...|...|...|...|...|...|

Specifically, this function extracts the necessary columns from the raw data file ("Sample", "Date", and those with "(Int)" in the column name), then joins this file and the project information file in a single dataframe by their shared column "Sample". 

This dataframe will further be used in the function `convertxrf()`, where the values are converted from kcps to µM. 

### convertxrf()

Once you have your dataframe, you go on to perform the actual calculations to convert your values from kcps to µM, ending up with another dataframe. For this function to work you will need to define several arguments:

* **setuppath**: This is the name of your "setup" file. It must be an Excel file, either .xlsx or .xls. 
* **year**: This is the year in which the crystal drift was measured closest to when your samples were measured. 
* **first_element**: This is the name of the first column in your dataframe that was created with the function `importxrf()` that is an element. For example, in the example above, there are the columns "Sample" and "Date.x" first, then the column named "C", so this is what we include in the function. 
* **last_element**: This is the name of the last column in your dataframe that is an element. In our example, this was the column "As". 

Here is a possible input:

```{r convertxrf, eval = FALSE}

df <- convertxrf(setuppath = "xrf_setup.xlsx", year = "2019", first_element = "C", last_element = "As")

```

This code will give a long dataframe, and you can see the first 96 rows [here](https://github.com/agryt/xrfr/blob/master/inst/extdata/project_calculated_sliced.csv). The table below shows the first few rows of this file:

|Sample|Date.x|Filter_type|...|Day|Treatment|Element|Concentration|Detection_limit|
|:----|:-----|:----------|:---|:--|:--------|:------|:------------|:--------------|
|COM001|04.12.2019 10:16:35|PC|...|0|K1|C|-4.19727307987776|1.22193802149026|
|COM001|04.12.2019 10:16:35|PC|...|0|K1|N|0.292013243234358|0.312338639119395|
|COM001|04.12.2019 10:16:35|PC|...|0|K1|O|0.328051062623146|0.0322146560280234|
|...|...|...|...|...|...|...|...|...|

Once you have this dataframe you need to save it. This is done by using the `write_csv()` function from the readr package (part of tidyverse). This is the file that you will use when performing any statistical analysis on the data or making plots. It is also necessary to save it if you with to use any of the functions in the xrfr package that further treats your dataset. 

### widen()

The file created using `convertxrf()` is good for further work in R/Rstudio, but if you want to look at your data it is not ideal. This function will show your data as is, only excluding columns you don't need and turning the dataframe from a long to a wide format. For this function to work you will need to define this argument:

* **projectpath**: This is the name of your "project file", the one you saved after having used `convertxrf()`. This must be a CSV file. 

A possible input could be this:

```{r widen, eval = FALSE}

df <- widen(projectpath = "project_longform.csv")

```

Running this code will give you a dataframe that looks something like this (LINK):

|Sample|...|Day|Treatment|C|N|O|Na|Mg|...|
|:----|:---|:--|:-------|:-|:-|:-|:-|:-|:-|
|COM001|...|0|K1|-4.19727308|0.29201324|0.32805106|-0.05557942|0.0353942597|...|
|COM002|...|0|K2|-21.69354177|0.14490883|0.90309406|-0.01586340|0.1147511764|...|
|COM003|...|0|K3|0.02039932|0.46326913|0.17582051|-0.05938854|0.0189933824|...|
|...|...|...|...|...|...|...|...|...|...|...|
